Emory Cardiovscular Biobank

### 2021-06-15 Manuscript/05-Discussion.Rmd
Adding Content

>>>>

## Discussion

In this chapter, we propose the spike-and-slab generalized additive models for high-dimensional analysis.Two fast computing algorithms, EM-IRLS and EM-coordinate descent are developed for expedited model fitting. EM-IRLS provides uncertainty measures, while EM-coordinate descent algorithm are extremely fast for model fitting. In the simulation studies, we shows the proposed model outperforms the state of the art GAM model, `mgcv` in prediction and function inference. Moreover, due to the model selection nature of the spike-and-slab prior, the proposed model can identify predictors that have large effects on the outcome. In addition, the model can be used to answer the question if non-linear effect are necessary.

In the manuscript, we describe the implementation of ssGAM-NM using EM-IRLS. Nevertheless, EM-IRLS are not limited to fitting ssGAM-NM. Other spike-and-slab prior can be fitted with EM-IRLS as well, e.g. mixture t-distribution, and mixture double exponential distributions. @Yi2012 demonstrated that t prior and double exponential prior can be seen hierarchical normal priors,
$$
\begin{aligned}
\beta_i \sim N(0, \tau^2_i),\\
\tau^2_i \sim Inv-\chi^2(v, s^2_j) \text{ or } \tau^2_i \sim Gamma(1, s^2/2)
\end{aligned}
$$
with $\tau^2$ integrated out analytically. Hence, the EM-IRLS can be easily applied to those two additional mixture priors. One of the advantage of using mixture $t$ prior over mixture normal prior is that $t$ distributions with low degree of freedom, e.g. $v=1$, have heavier tail than the normal distribution, which is also a special case of $t$ distribution when $v=\infty$. The heavier tails help to obtain posteriors that concentrated around the true nonzero coefficients at optimal rates. [@Rockova2014a]

### 2021-06-15 Manuscript/SS_GAM.tex
Adding Content

\hypertarget{simulation}{%
\section{Simulation}\label{simulation}}

@@

>>>>

\hypertarget{simulation-study}{%
\subsection{Simulation Study}\label{simulation-study}}

In this section, we compare the performance of the proposed models to 3
other models: COSSO, adaptive COSSO ,mgcv. COSSO is xxx method. Adaptive
COSSO is another method, mgcv is another method. To demonstrate the
correctness and feasibility of the proposed SS-GAM, we follow the
logistic regression simulation introduced in Bai (Work in Progress). We
first generate \(n=500\) training data sample with number of predictors
\(p=4, 10, 50, 200\) respectively, where the predictors
\(X \sim MVN_{n\times p}(0, I_{P})\). We simulate the outcome data from
three distributions, gaussian, poisson, and binomial with the identity
link, logit link \(g(x) = \log(\frac{x}{1-x})\), and log link
\(g(x) = log(x)\) functions to mimic the common use of generalized
model. The outcome is simulated via the following function \[
\mathbb{E}(Y) = g^{-1}(5 \sin(2\pi x_1) - 4 \cos(2\pi x_2 -0.5) + 6(x_3-0.5) - 5(x_4^2 -0.3))
\] for gaussian and binomial outcome and shrinked effects

\[
\mathbb{E}(Y) = g^{-1}(\sin(2\pi x_1) - 2\cos(2\pi x_2 -0.5) + 0.4(x_3-0.5) - 2(x_4^2 -0.3))
\] for stable poisson outcome. For the Gaussian outcome where dispersion
are required, the dispersion parameter was set to 1. We follow the same
data generating functions in Bai. In this data generating process, we
have \(x_1, x_2, x_3, x_4\) as the active covariates, while the rest
covariates are inactive \(f_j(x_j) = 0\) for \(j = 4, \dots, p\).
Another set of independent sample of size
\(n_{test}=1000\),(\(x_{new} ,y_{new}\)), are created following the same
data generating process, serving as the testing data. For mgcv and
proposed models where choice of spline function are customizable, we
choose to cubic regression spline with 10 basis. In COSSO and adaptive
COSSO where the choice of spline are pre-determined, we set to have 10
bases. The corresponding number of parameter that we estimate with the
model is 37, 82, 451 and 1801. We are run the simulation where the
number of bases are increased to 25 deliberately. To evaluate the
performance of the models, mean squared error, mean absolute error for
gaussian model, AUC and misclassifcation rate are calculated for
binomial model, and deviance are calculated for Poisson model, averaging
across50 simulations. The results are tabulated in Table 1, 2, 3.

\%are calculated to assess out-of-sample estimation accuracy,

\%We also calculated the area under the curve (AUC) of the receiver
operating characteristic curve for \(y_{new}\). We will also visually
display the estimated smooth function with the smooth function of truth
to compare if the models can infer the smooth function correctly,
despite the prediction accuracy.

\%When constructing the smooth function of each variables, we constructs
the natural cubic spline with degree of freedom 25. In other words, the
design matrix of each smooth function will have 26 columns. The total
number of columns for \(p=4, 10, 50\) is \(105, 261, 1301\)
respectively, where intercept is kept in the model when fitting. The
\(p=50\) can mimic the high dimensional setting as the design matrix are
in high-dimensional setting (\(p >> n\))

\%We benchmark the performance of the proposed methods with the state of
art GAM package \texttt{mgcv}{[}TODO: add citation{]}. In the future
steps, we will compare our proposed model with other high dimensional
GAM applications.

\hypertarget{simulation-results}{%
\subsubsection{Simulation Results}\label{simulation-results}}

Overall, the proposed priors outperforms the state-of-art GAM model
\texttt{mgcv} when the dimensionality is low, median, and high. The
proposed priors predicts better than \texttt{mgcv} and infers better of
the underlying smooth function. Most importantly, \texttt{mgcv} is
infeasible in the high dimensional setting, i.e.~when \(p=50\) where the
number of parameters are greater than the sample size.

\hypertarget{prediction-performace}{%
\paragraph{Prediction Performace}\label{prediction-performace}}

The prediction performance is tabulated in the table below for AUC and
Supplementary tables for MSE and misclassification rate. In summary, the
proposed Bayesian models outperforms the benchmark model \texttt{mgcv}
regardless the choice of prior or fitting algorithm. The models with
IRLS fitting algorithm give slightly better performance than the the
model with cyclic coordinate descent algorithm. The advantage diminishes
in the high dimensional setting (p=50). This is possible because the SS
normal spline prior provides a more smooth solution, which would work
well in a low to median dimension setting. However, in the high
dimensional setting, there are not enough sparsity in the model, i.e.~SS
normal prior will provides a more complex model than necessary. This is
confirmed with the visualization in the following section.

\hypertarget{visualization}{%
\paragraph{Visualization}\label{visualization}}

We plot the estimated smooth functions of the four active variables for
the 100 simulation iterations, along with the `true' smooth function.
Across different settings of predictors, the patterns look similar.
First of all, in all the settings, \texttt{mgcv} tends to provide more
extreme models, where in the graphs the smooth function have much higher
bound than the true smooth function. Rarely the estimated smooth
function from \texttt{mgcv} matches with the true smooth function. In
comparison, the proposed method can interpolate the shape of the true
smooth function. However, different priors and fitting algorithms have
different advantages. For example, in Figure 1, we see that the SS
normal distribution have problem with the tail of the smooth function,
especially obvious in the estimations of the variable \emph{x4}.
However, it would have less sparse solutions, which is more likely to
accurately interpolate the smooth function especially when the shape is
complicated (see variable \emph{x2}). On the contrary, the SS double
exponential prior model encourages sparse solutions, which can sometime
omit active predictors.

@@

\hypertarget{discussion-1}{%
\subsection{Discussion}\label{discussion-1}}

In this chapter, we propose the spike-and-slab generalized additive
models for high-dimensional analysis.Two fast computing algorithms,
EM-IRLS and EM-coordinate descent are developed for expedited model
fitting. EM-IRLS provides uncertainty measures, while EM-coordinate
descent algorithm are extremely fast for model fitting. In the
simulation studies, we shows the proposed model outperforms the state of
the art GAM model, \texttt{mgcv} in prediction and function inference.
Moreover, due to the model selection nature of the spike-and-slab prior,
the proposed model can identify predictors that have large effects on
the outcome. In addition, the model can be used to answer the question
if non-linear effect are necessary.

In the manuscript, we describe the implementation of ssGAM-NM using
EM-IRLS. Nevertheless, EM-IRLS are not limited to fitting ssGAM-NM.
Other spike-and-slab prior can be fitted with EM-IRLS as well,
e.g.~mixture t-distribution, and mixture double exponential
distributions. \citet{Yi2012} demonstrated that t prior and double
exponential prior can be seen hierarchical normal priors, \[
\begin{aligned}
\beta_i \sim N(0, \tau^2_i),\\
\tau^2_i \sim Inv-\chi^2(v, s^2_j) \text{ or } \tau^2_i \sim Gamma(1, s^2/2)
\end{aligned}
\] with \(\tau^2\) integrated out analytically. Hence, the EM-IRLS can
be easily applied to those two additional mixture priors. One of the
advantage of using mixture \(t\) prior over mixture normal prior is that
\(t\) distributions with low degree of freedom, e.g.~\(v=1\), have
heavier tail than the normal distribution, which is also a special case
of \(t\) distribution when \(v=\infty\). The heavier tails help to
obtain posteriors that concentrated around the true nonzero coefficients
at optimal rates. \citep{Rockova2014a}

### 2021-06-15 Real_Data/Emory_Card_Biobank/Report.Rmd
Adding Content

library(targets)
library(readr)
library(here)
library(gtsummary)
library(tidyverse)

@@

@@

lasso_mdl <- glmNet(x = train_data %>% select(starts_with("mz")) %>% data.matrix,
                    y = train_data %>% pull(death3yr), alpha = 1,
                    family = "binomial",lambda = lasso_mdl_cv$lambda.1se)

@@

measure.bh(lasso_mdl, new.x = test_data %>% select(starts_with("mz")) %>% data.matrix,
           new.y = test_data %>% pull(death3yr))
```

```{r}
library(BHAM)
library(BhGLM)

mz_names <- train_dat %>% select(starts_with("mz")) %>% names  
sm_df <- data.frame(
  Var = mz_names,
 Func = "s",
 Args ="bs='cr', k=5"
)

@@

mdl_defaul <- bmlasso(dsn_mat, train_data$death3yr, family = "binomial")

mdl_dflt <- bmlasso_spline(dsn_mat, train_data$death3yr, family = "binomial",
                           group = make_group(names(dsn_mat)))

# Training Data Performance
measure.bh(mdl_defaul)
measure.bh(mdl_dflt)

test_sm_dat <- BHAM::make_predict_dat(sm_obj$Smooth, dat=test_data)

# Testing Data Performance
measure.bh(mdl_defaul, new.x = test_sm_dat, new.y=test_data$death3yr)
measure.bh(mdl_dflt, new.x = test_sm_dat, new.y=test_data$death3yr)
```

>>>>

library(targets)
library(readr)
library(here)
library(gtsummary)
library(tidyverse)
library(BhGLM)

@@

## Screening
### Linear Effect Screening
```{r}
# feature_name <- 
linear_screening_res <- train_dat %>% select(starts_with("mz")) %>% 
  names() %>% 
map_dfr(#feature_name[1:2],
        .f = function(name, .dat){
  # name <- feature_name[1]
  y <- .dat %>% pull(death3yr)
  x <- .dat %>% pull({{name}})
  glm(y ~ x, family = binomial()) %>% 
    tidy() %>% 
    filter(term == "x") %>% 
    select(p.value) %>% 
    mutate(var = name,
           p.value)

},
.dat = train_dat)

lnr_scn_res <- data.frame(
linear_screening_res,
p.adj = p.adjust(linear_screening_res$p.value, "fdr")
) %>% arrange(p.adj) %>% filter(p.adj<0.2) #%>% 
  # slice_head(n = 100)
```

### GAM Screening
```{r}

library(mgcv)

# feature_name <- 
gam_screening_res <- train_dat %>% select(starts_with("mz")) %>% 
  names() %>% 
map_dfr(#feature_name[1:2],
        .f = function(name, .dat){
  # name <- feature_name[1]
  y <- .dat %>% pull(death3yr)
  x <- .dat %>% pull({{name}})
  mgcv::gam(y ~ s(x, bs = "cr", k = 5), family = binomial(),
            data = data.frame(x, y)) %>% 
    tidy() %>% 
    filter(term == "s(x)") %>% 
    select(p.value) %>% 
    mutate(var = name,
           p.value)

},
.dat = train_dat)

gam_scn_res <- data.frame(
gam_screening_res,
p.adj = p.adjust(gam_screening_res$p.value, "fdr")
) %>% arrange(p.adj) %>% #filter(p.adj<0.2)
  head(200)
```

@@

lasso_mdl <- glmNet(x = train_data %>% select(starts_with("mz")) %>% data.matrix,
                    y = train_data %>% pull(death3yr), alpha = 1,
                    family = "binomial",lambda = lasso_mdl_cv$lambda.min)

@@

measure.bh(lasso_mdl, new.x = test_data %>% select(starts_with("mz")) %>% as.matrix,
           new.y = test_data %>% pull(death3yr))

lasso_mdl <- glmnet(x = train_data %>% select(starts_with("mz")) %>% as.matrix,
                    y = train_data %>% pull(death3yr), alpha = 1,
                    family = "binomial",lambda = lasso_mdl_cv$lambda.min)

assess.glmnet(lasso_mdl, newx = train_data %>% select(starts_with("mz")) %>% as.matrix, newy = train_data %>% pull(death3yr))

assess.glmnet(lasso_mdl, newx = test_data %>% select(starts_with("mz")) %>% as.matrix, newy = test_data %>% pull(death3yr))

```

lasso_mdl_cv <- cv.glmnet(x = train_data %>% select(gam_scn_res %>% pull(var)) %>% data.matrix,
                    y = train_data %>% pull(death3yr), alpha = 1,
                    family = "binomial", type.measure = "auc")

lasso_mdl <- glmNet(x = train_data %>% select(gam_scn_res %>% pull(var)) %>% data.matrix,
                    y = train_data %>% pull(death3yr), alpha = 1,
                    family = "binomial",lambda = lasso_mdl_cv$lambda.min)

measure.bh(lasso_mdl, new.x = test_data %>% select(gam_scn_res %>% pull(var)) %>% as.matrix,
           new.y = test_data %>% pull(death3yr))

lasso_mdl <- glmnet(x = train_data %>% select(starts_with("mz")) %>% as.matrix,
                    y = train_data %>% pull(death3yr), alpha = 1,
                    family = "binomial",lambda = lasso_mdl_cv$lambda.min)

assess.glmnet(lasso_mdl, newx = train_data %>% select(starts_with("mz")) %>% as.matrix, newy = train_data %>% pull(death3yr))

assess.glmnet(lasso_mdl, newx = test_data %>% select(starts_with("mz")) %>% as.matrix, newy = test_data %>% pull(death3yr))

```

```{r}
library(BHAM)

mz_names <- train_dat %>% select(gam_scn_res %>% pull(var)) %>% names  
sm_df <- data.frame(
  Var = mz_names,
 Func = "s",
 Args ="bs='cr', k=5"
)

@@

mdl_defaul <- bmlasso(dsn_mat, train_data$death3yr, family = "binomial", ss  = c(lasso_mdl$prior.scale, 0.5))

mdl_dflt <- bmlasso_spline(dsn_mat, train_data$death3yr, family = "binomial",
                           group = make_group(names(dsn_mat)),
                           ss  = c(lasso_mdl$prior.scale, 0.5))

# Training Data Performance
cat("bmlasso model\n")
measure.bh(mdl_defaul)
cat("bmlasso_spline model\n")
measure.bh(mdl_dflt)

```

```{r}
test_sm_dat <- BHAM::make_predict_dat(sm_obj$Smooth, dat=test_data)

# Testing Data Performance
cat("bmlasso model\n")
measure.bh(mdl_defaul, new.x = test_sm_dat, new.y=test_data$death3yr)
cat("bmlasso_spline model\n")
measure.bh(mdl_dflt, new.x = test_sm_dat, new.y=test_data$death3yr)
```
